Data Specialization for a Staged Lambda Calculus

### 2015-02-05 paper/splitting.tex
edits to splitting section

Recall that partial values are multistage terms with the character of a value at stage~\bbone\ and of a term at stage~\bbtwo.

For example, the term
\begin{lstlisting}
1`(1+2, next{`2`7 + 5`1`})`
\end{lstlisting}
evaluates via $\redonesym$ to:
\begin{lstlisting}
2`[yhat|->7+5]` 1`(3,next{`2`yhat`1`})`
\end{lstlisting}

What we desire is a simple way to translate this partial value into a form that separates the stage~\bbone\ content (the value \texttt{3}) 
from the stage~\bbtwo\ content (the term $\letin \yhat {\mathtt{7+5}} \yhat$),
allowing us to later define a term translation that respects this value translation.

For this purpose, we define a relation called {\em masking}, written $\vsplito$, 
which sends partial values to {\em masked values}.
Masked values have only one form, written $\mval {i} {r}$, 
where $i$ is a monostage value (the {\em immediate value}) and $r$ is a monostage term (the {\em resumer}).
The masking relation is defined in \ref{fig:valMask}.

@@

The example above is masked into,
\begin{lstlisting}
mv((3,()), let yhat=7+5 in ((),yhat))
\end{lstlisting}

\subsection{Stage \bbone\ Term Splitting}
\label{sec:split-one}

Recall that a term $\colone e A$ represents a multistage computation with a multistage result.
To represent this in a separated form, we use the ``multi-output pipeline'' construct $\pipeM c l r$, 
where $c$ is a monostage term and $r$ is a monostage term open on the variable $l$.
This can be partially evaluated (into a masked value) with the following rule:
\[
\infer {\sepredone {\pipeM c l r} {\mval i {[b/l]r}}} {\reduce c {(i,b)}}
\]
We require that $c$ (the \emph{combined term}) reduce to the tuple $(i,b)$,
where $i$ becomes the immediate result of the masked value,
and where $b$ is plugged into the open variable of the resumer.
We refer to $b$ as the \emph{boundary value} of the split since it carries information between stage~\bbone\ and stage~\bbtwo\ subcomputations.

The splitting algorithm for stage~\bbone\ terms, as specified by the
judgment $e \splitonesym \pipeM c l r$ in \cref{fig:termSplit}, proceeds
recursively on the structure of~$e$.
Our goal for the transformation is that 
if $\diaone e {\xi,v}$ and $[\xi,v] \vsplito \mval i r$, 
then $\splitone e A {\pipeM c l {r'}}$ and $\sepredone {\pipeM c l {r'}} {\mval i {r''}}$,
where $r \equiv r''$ in some standard equivalence.
Intuitively, translation and evaluation should commute.

@@

For all non-terminals (except $\next$),
splitting descends into $e$, recursively splitting its $n$ subterms
to produce their respective combined terms $c_1,\ldots,c_n$ and resumers $r_1, \ldots, r_n$.
The combined term of $e$ is formed by binding $c_1,\ldots,c_n$
to the patterns $(y_1,z_1),\ldots,(y_n,z_n)$
to isolate \bbone-results from boundary values. Then,
the \bbone-result of $e$ is formed by replacing $e$'s subterms with $y_1,\ldots,y_n$.
The resumer binds the boundary values $b_1,\ldots,b_n$ to an
argument $(l_1,\ldots,l_n)$ in a term that has the same structure
of~$e$ but where each subterm is replaced by its resumer ($r_i$'s).

Splitting {\tt case} yields a combined term that executes one of the branches' combined terms based on the \bbone-result $y_1$ of the predicate.
The boundary value $b_i$ for this branch is injected and bundled with that of the predicate ($b_1$).   
$b_i$ is cased in the resumer to determine which branch's resumer should be executed.
{\tt case} and $\pause$ are the only two rules where splitting adds non-trivial logic is added to the precomputation.

@@

Since the results of splitting $\next$ terms depend on the output of splitting its stage~\bbtwo\ subterm,
we defer description of $\next$ until after describing stage~\bbtwo\ term splitting.

\subsection{Stage \bbtwo\ Term Splitting}

Because stage~\bbtwo\ terms in \lang\ fully reduce to values (as opposed to partial values),
stage~\bbtwo\ term splitting assumes a simpler form than does stage~\bbone\ term splitting. 
Specifically, splitting a term $\coltwo e A$ in \lang\ generates a ``single-output pipeline''
term $\pipeS p l r$ where $p$ is a monostage {\em precomputation} term 
that performs all of the stage~\bbone\ subcomputations in $e$, and $l.r$ 
is a resumer which that performs all of the stage~\bbtwo\ subcomputations in $e$.
This can be partially evaluated into a residual via the rule,
\[
\infer {\sepredone {\pipeS p l r} {[b/l]r}} {\reduce p b} 
\]
Note that in contrast to the combined term produced during stage~\bbone\ splitting,
the precomputation $p$ generated by stage~\bbtwo\ splitting exists entirely
to compute the boundary value expected by the resumer.

% \ur{I don't quite understand the following.}
% (Thus, the resumer, bound to the value output by $p$, is equivalent to
% the residual term defined in~\ref{ssec:dynamics})

As with stage~\bbone\ terms, the splitting algorithm for stage~\bbtwo\ terms,
as specified by the judgment $e \splittwosym [p,l.r]$ in \cref{fig:termSplit}, is defined on the local structure of $e$.
In the terminal cases of
constants and variables, splitting generates trivial precomputations that are \texttt{()}, and resumers consisting of the original term.
For example, the integer constant \texttt{3} splits into the
precomputation \texttt{()} and resumer \texttt{\_=>3}.

For all non-terminal terms~$e$ in \lang\ (except $\prev$
and $\pause$)
% recursively splitting
% subterms and generating precomputations $p_1, \ldots, p_n$ and
% resumers $r_1, \ldots, r_n$ for a term with~$n$ immediate subterms.
the precomputation of~$e$ is defined as the combined precomputations of $e$'s $n$ subterms:
$p=(p_1,\ldots,p_n)$.  The resumer binds each boundary value to an
argument $(l_1,\ldots,l_n)$ in a term that has the same structure
of~$e$ but where each subterm is replaced by its corresponding resumer $r_i$.

A notable properly of splitting stage~\bbtwo\ \texttt{case}s and functions is that the
precomputation of subterms is lifted out from underneath stage
\bbtwo\
binders.  % TODO: should probably draw a parallel to the same behavior in dynamics

Splitting $\prev$ generates a precomputation that projects the \bbone-result of its stage~\bbone\ subterm.
Since the argument to $\prev$ is of $\fut$ type, its \bbone-result reduces to \texttt{()}, justifying why it can be thrown away.
$\pause$ treats the entire combined term of its stage~\bbone\ subexpression as a precomputation, 
and projects out the integer result in the resumer. 
\footnote{The resumer of an integer expression is usually trivial, 
but we have to include it here for termination purposes.} 
Finally, splitting $\next$ simply tuples up the precomputation of its stage~\bbtwo\ subterm with a trivial \bbone-result \texttt{()}.

>>>>

The point of the masking operation is simply to convert a partial value into a masked value $\mval i q$
by assigning all of the stage \bbone\ content to $i$ and all of the stage \bbtwo\ content to $q$.
The rules of the masking relation are given in \ref{fig:valMask}.

@@

\subsection{Term Splitting at \bbonem}
\label{sec:split-one}

We now show how to translate terms $\colone e A$ into the form $\pipeM c l r$,
pursuant to the correctness condition given in \ref{fig:splittingSummary}.
The algorithm is specified by the $\splitonesym$ relation (\cref{fig:termSplit}), 
which proceeds recursively on the structure of~$e$.

@@

For all non-terminals (except \texttt{next}),
splitting descends into $e$, recursively splitting its $n$ subterms
to produce their respective combined terms $c_1,\ldots,c_n$ and resumers $r_1, \ldots, r_n$.
The combined term of $e$ is formed by binding $c_1,\ldots,c_n$
to the patterns $(y_1,z_1),\ldots,(y_n,z_n)$
to isolate \bbone-results from boundary values. Then,
the \bbone-result of $e$ is formed by replacing $e$'s subterms with $y_1,\ldots,y_n$.
The resumer binds the boundary values $b_1,\ldots,b_n$ to an
argument $(l_1,\ldots,l_n)$ in a term that has the same structure
of~$e$ but where each subterm is replaced by its resumer ($r_i$'s).

Splitting {\tt case} yields a combined term that executes one of the branches' combined terms based on the \bbone-result $y_1$ of the predicate.
The boundary value $b_i$ for this branch is injected and bundled with that of the predicate ($b_1$).   
$b_i$ is cased in the resumer to determine which branch's resumer should be executed.
{\tt case} and \texttt{hold} are the only two rules where splitting adds non-trivial logic is added to the precomputation.

@@

Since the results of splitting \texttt{next} terms depend on the output of splitting its stage~\bbtwo\ subterm,
we defer description of \texttt{next} until after describing stage~\bbtwo\ term splitting.

\subsection{Term Splitting at \bbtwo}

Because stage~\bbtwo\ terms in \lang\ reduce to monostage residuals (as opposed to partial values),
term splitting at \bbtwo\ assumes a simpler form than the version at \bbonem\ does. 
The algorithm is specified by the $\splittwosym$ relation in \cref{fig:termSplit}.

In the terminal cases of
constants and variables, splitting generates trivial precomputations that are \texttt{()}, and resumers consisting of the original term.
For example, the integer constant \texttt{3} splits into the
precomputation \texttt{()} and resumer \texttt{\_=>3}.

More generally, for all (except \texttt{prev} and \texttt{hold}) 
$n$-ary terms $e = \mathcal{C}\ttlpar e_1 \ttsemi \ldots \ttsemi e_n \ttrpar$ 
the precomputation is the tupled precomputations of $e$'s $n$ subterms:
$p=(p_1,\ldots,p_n)$.  The resumer binds each boundary value to an
argument $(l_1,\ldots,l_n)$ in a term that has the same structure
of~$e$ but where each subterm is replaced by its corresponding resumer:
$r = \mathcal{C}\ttlpar r_1 \ttsemi \ldots \ttsemi r_n \ttrpar$ .
Notably, at \texttt{case}s and functions the
precomputation of subterms is lifted out from underneath stage \bbtwo\ binders.  
% TODO: should probably draw a parallel to the same behavior in dynamics

Splitting \texttt{prev} generates a precomputation that projects the \bbone-result of its stage~\bbone\ subterm.
Since the argument to \texttt{prev} is of $\fut$ type, its \bbone-result reduces to \texttt{()}, justifying why it can be thrown away.
\texttt{hold} treats the entire combined term of its stage~\bbone\ subexpression as a precomputation, 
and projects out the integer result in the resumer. 
\footnote{The resumer of an integer expression is usually trivial, 
but we have to include it here for termination purposes.} 
Finally, splitting \texttt{next} simply tuples up the precomputation of its stage~\bbtwo\ subterm with a trivial \bbone-result \texttt{()}.
